---
title: Homework 2
subtitle: Pandas Basics
date: 2024-02-13
execute: 
  echo: false
  eval: false
  warning: false
  message: false
from: markdown+emoji
comments: false
---

```{r}
#| include: false

library(tidyverse)
library(DT)

reticulate::use_condaenv("/Users/bchoe/anaconda3", required = TRUE)
```


# Direction


- Please submit your Jupyter Notebook for Homework 2 to the Brightspace with the name below:

  - `danl-m1-hw2-LASTNAME-FIRSTNAME.ipynb`\
  ( e.g., `danl-m2-hw1-choe-byeonghak.ipynb` )

- The due is February 20, 2024, 7:00 P.M.

- Please send Byeong-Hak an email (`bchoe@geneseo.edu`) if you have any questions.

-   Write a Python Notebook (`*.ipynb`) to answer each question.

-   Make at least some simple comment (`# ...`) in each question.

-   Make one text cell to explain things in each question.

# Load Libraries

Import the `pandas` library.

```{python}
#| echo: true
import pandas as pd
```

Below is `nhl` `DataFrame` that reads the file `NHL-1617.csv` containing data of all non-goalie NHL players who had at least some _ice time_, i.e. playing time, in the 2016/2017 regular season - without play-offs. 

```{python}
#| echo: true
nhl = pd.read_csv('https://bcdanl.github.io/data/NHL-1617.csv')
```

<div style="position: relative;">
  <button onclick="scrollTable(-200)" style="position: absolute; z-index: 1000; top: 0; left: 70px; border: 1px solid grey; width: 25px; height: 41px; background-color: #ededed;">←</button>
  <button onclick="scrollTable(200)" style="position: absolute; z-index: 1000; top: 0; left: 95px; border: 1px solid grey; width: 25px; height: 41px; background-color: #ededed;">→</button>
</div>


<script>
function scrollTable(offset) {
  // Ensure we're targeting the scroll body of the DataTable specifically
  var scrollBody = $('#exampleTable .dataTables_scrollBody');
  
  // Check if the scrollBody is found
  console.log(scrollBody.length); // For debugging, should be 1 if found
  
  scrollBody.animate({
    scrollLeft: '+=' + offset
  }, 'fast');
}
</script>


```{r}
#| eval: true
#| echo: false
DT:::datatable(readr::read_csv('https://bcdanl.github.io/data/NHL-1617.csv'), 
          extensions = c('Buttons','FixedColumns'),
          options = list(
            dom = 'Bfrtip',
            buttons = c('csv'),
            scrollX = TRUE,
            fixedColumns = list(leftColumns = 2)
            ), 
          elementId = "exampleTable"
          )
```

## Variable Description
- `id_player`: Player ID
- `Born`: Date of Birth
- `City`: City of Birth
- `Cntry`: Country of Birth
- `Nat`: Nationality
- `Last_Name`: Last name
- `First_name`: First Name
- `Position`: Position
- `Team`: Team
- `GP`: The number of games
- `G`: The number of goals
- `A`: The number of assists
- `TOI`: The total time on ice (in second)
- `TOI_GP`: Average amount of playing time per game (in minute)

<br><br>

## Question 1

How many players had at least some ice time?

*Answer*

```{python}

```


<br><br>


## Question 2

Who is the top scorer in terms of points?
*Answer*
```{python}

```

<br><br>

## Question 3
How many Russian (non-goalie) players had some ice time in there 2016/2017 regular season? (Hint: Russians are indicated by "RUS".)

*Answer*
```{python}

```


<br><br>


## Question 4
Create a `DataFrame` named `nhl_RUS` that includes only observations for Russian players, retaining just their `Last_Name` and `First_Name`.

*Answer*
```{python}

```

<br><br>

## Question 5
Who performed best among the Russian players in terms of `G`? How many goals did he have?

*Answer*
```{python}

```

<br><br>


## Question 6
Which three nationalities have the highest number of players?

*Answer*
```{python}

```

<br><br>
