---
title: Classwork 2
subtitle: Pandas Basics [Classwork 2 may be subject to change during the Module.]
date: 2024-02-13
execute: 
  eval: false
  echo: false
  warning: false
  message: false
  
from: markdown+emoji
comments: false
---

- The following is the Python libraries we need for this homework.
```{python}
#| echo: true
import pandas as pd
import seaborn as sns
```



# Question 1

Load the CSV file, `beer_markets_cleaned.csv`, from the class website:
```{python}
#| echo: true
beer_markets = pd.read_csv('https://bcdanl.github.io/data/beer_markets_cleaned.csv')
```


- Description of DataFrame `beer_markets` 
  - One single observation (row) corresponds to information regarding one household and its beer purchase.
  - DataFrame `beer_markets` does not have any missing values.

<br>

- Variable (column) description
  - `hh`: an identifier of the household;
  - `purchase_desc`: details on the purchased item;
  - `quantity`: the number of items purchased;
  - `brand`: Bud Light, Busch Light, Coors Light, Miller Lite, or Natural Light;
  - `spent`: total dollar value of purchase;
  - `beer_floz`: total volume of beer, in fluid ounces;
  - `price_per_floz`: price per fl.oz. (i.e., beer spent/beer floz);
  - `container`: the type of container;
  - `promo`: Whether the item was promoted (coupon or otherwise);
  - `market`: Scan-track market (or state if rural);
  - demographic data, including gender, marital status, household income, class of work, race, education, age, the size of household, and whether or not the household has a microwave or a dishwasher.

<br><br>

## Q1a
- Sort the DataFrame `beer_markets` by `hh` in ascending order.


<br>

*Answer*

```{python}
```

<br><br>

## Q1b
Count the number of households for each market.
<br>

*Answer*

```{python}

```

<br><br>

## Q1c

Find the top 5 beer markets in terms of the number of households that purchased beer.

<br>

*Answer*

```{python}

```

```{r}

```

<br><br>

## Q1d
Sum of `beer_floz` for each `market`.
<br>

*Answer*

```{python}

```

<br><br>

## Q1e
Find the top 5 beer markets in terms of the amount of total beer consumption.
<br>

*Answer*

```{python}

```


```{r}

```

<br><br>

## Q1f
- Variable `price_per_floz` is continuous.
- Variable `brand` is categorical.

Describe the distribution of `price_per_floz` for each `brand` using `seaborn`.

Make a simple comment on comparison for the distribution of `price_per_floz` across brands.
<br>

*Answer*

```{python}

```

```{r}

```

<br>

- `BUD LIGHT`, `COORS LIGHT`, and `MILLER LITE` have the similar distribution of `price_per_floz` with each other.

- `BUSCH LIGHT` and `NATURAL LIGHT` have the similar distribution of `price_per_floz` with each other.

- Overall, `BUD LIGHT`, `COORS LIGHT`, and `MILLER LITE` are more expensive than `BUSCH LIGHT` and `NATURAL LIGHT`.



<br><br>

## Q1g

- Both variables `price_per_floz` and `beer_floz` are continuous.

- Describe the relationship between `price_per_floz` and `beer_floz` by brand using `seaborn`.

- Make a simple comment on the visualization result regarding how the relationship between `price_per_floz` and `beer_floz` varies by `brand`.
<br>

*Answer*

```{python}

```

```{r}

```



